---
title: "ARCS_Part3_intructor"
format: html
editor: visual
---

# Time for some coding

## Load packages used for this session using Pacman()

The **`pacman`** package helps ensure your script is **reproducible** as it automatically **installs and loads** the packages on someone else's computer that are needed to run your code. You can also add to the list if more packages are needed.

```{r}

if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  magrittr,     
  here,
  readr,
  janitor
)
```

## Importing data into R using the **'here()'** and **'readr()'** packages

The **`here`** package improves reproducibility and sharing by creating file paths relative to your **R project's root directory**, eliminating the need for hard-coded absolute paths - as long as your script file is saved in the same folder as your R project

### Run this code to make sure you are currently in the correct R Project

```{r}

here()
```

### OPTIONAL: If you are not in the current project you can run the following:

```{r}

## Replace 'path/to/your/desired/folder' with the actual path here:

#set_here("path/to/your/desired/folder")
```

## Let's import the two data files for this instruction

```{r}

## Add the file name for the data we want including the file type (.csv)

# This will import the raw data file
raw_data <- read_csv(here("raw_data", "GSE113754_GeneLevel_Raw_data_3.csv"))

# this will import the meta data file
raw_metadata <- read_csv(here("raw_data", "GSE113754_filtered_metadata.csv"))

## Used to show the difference between a tibble and regular DF (read.csv)
#raw_data_b <- read.csv(here("raw_data", "GSE113754_GeneLevel_Raw_data_3.csv"))
```

In this chunk we used *read_csv* instead of *read.csv*

1.  Part of the Tidyverse ecosystem, it has a clear system for defining column types, making scripts more reliable and reproducible
    -   For example, col_double instead of numeric/integer
2.  Reads larger data files faster
3.  Returns a *tibble,* a modern data frame with features that make it print nicely, displays on a screen better, does not change variables names, and does not use row labels
    -   Even the outputs for functions like str() are nicer looking and more informative

# Let's start with the metadata file

## Inspect data

### Physically look at the data you just imported

```{r}
# Add dataframe name to the function

view(raw_metadata)
```

### Use the head() function to see the first 6 rows of data, table dimension, column names, column data types, and row values

```{r}
# Add dataframe name to the function

head(raw_metadata)
```

### Use the str() function to see the structure, data types, dimensions, classes, and how data are organized 

Good for understanding the data structure

```{r}
# Add dataframe name to the function

str(raw_metadata)
```

### EXTRA: We can even find the unique values within a variable and their count

```{r}
raw_metadata %>% count(genotype)

## This code is similar to the unique() and table() exercise from last class
#meta_data %>% 
  #count(genotype) %>% 
  #distinct(genotype) %>% 
  #map(unique) %>% 
  #map(table)
```

## Clean raw_metadata

-   Start by creating a new dataset by assigning a new object using '\<-'

-   Separate each function the pipe operator (%\>%)

-   **TIP: you can run the code between each %\>% to see how the dataset changes and inspect any errors with your code**

### What changes do we want to make?

1.  Fix column names

2.  Rename characteristic to condition

3.  Clean up the values under condition

4.  Clean up the values under genotype

5.  Combine columns: condition and genotype (geno_cond)

6.  Keep the following in this order: sample_id, genotype, condition, geno_cond

7.  Make the values under sample_id lower case

### Step 1: Fix all column names

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() 
```

### Step 2: Rename:

-    column 1 (x1) to *sample_id*
-   *characteristic* to *condition*

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() %>% 
  rename(sample_id = x1, condition = characteristics)
```

### Step 3: Update values under condition

-   if value includes "homecage", then change to "normal"
-   if value includes "sleep deprivation", then change to "sleep_deprived"
-   NOTE: notice how we use grepl here like we did during Sofia's presentation, again we can use base R code within the Tidyverse

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() %>% 
  rename(sample_id = x1, condition = characteristics) %>% 
  mutate(
    condition = case_when(
      grepl("homecage", condition) ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition)) 
```

### Step 4: Update values under genotype

-   if value includes "WT", then change to "WT"
-   if value includes "shank3", then change to "shank3"

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() %>% 
  rename(sample_id = x1, condition = characteristics) %>% 
  mutate(
    condition = case_when(
      grepl("homecage", condition) ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition))  %>%
  mutate(
    genotype = case_when(
      grepl("WT", genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype))
```

### Step 5: Combine columns "condition" and "genotype" and create new variable with their values called "geno_cond"

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() %>% 
  rename(sample_id = x1, condition = characteristics) %>% 
  mutate(
    condition = case_when(
      grepl("homecage", condition) ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition))  %>%
  mutate(
    genotype = case_when(
      grepl("WT", genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype)) %>%
  mutate(
    geno_cond = paste0(genotype, condition)) 
```

### Step 6: Keep the following in this order: sample_id, genotype, condition, geno_cond

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() %>% 
  rename(sample_id = x1, condition = characteristics) %>% 
  mutate(
    condition = case_when(
      grepl("homecage", condition) ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    ))  %>%
  mutate(
    genotype = case_when(
      grepl("WT", genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype
    )) %>%
  mutate(
    geno_cond = paste0(genotype, condition)) %>% 
  select(sample_id, genotype, condition, geno_cond)
```

### Step 7: Make the values under sample_id lower case for processing in the next step (cleaning raw_data)

```{r}
clean_meta_data <- raw_metadata %>% 
  clean_names() %>% 
  rename(sample_id = x1, condition = characteristics) %>% 
  mutate(
    condition = case_when(
      grepl("homecage", condition) ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    ))  %>%
  mutate(
    genotype = case_when(
      grepl("WT", genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype
    )) %>%
  mutate(
    geno_cond = paste0(genotype, condition)) %>% 
  select(sample_id, genotype, condition, geno_cond) %>%
  mutate(sample_id = str_to_lower(sample_id))
```

# Now let's clean raw_data

## Inspect data

```{r}
view(raw_data)
```

```{r}
head(raw_data)
```

```{r}
str(raw_data)
```

### What issues did you find? (see tenet 8)

-   Column 1 has no label

-   Column 1 includes row numbers, is inserted in error, & shouldn't be there

-   Column 2 has no label

-   Column 3, "gene symbol", column has a space

-   All columns beginning with "GSM" are upper-case

### EXTRA: let's look for NA values 

Use filter(), if_any(), everything() functions with is.na to search all rows and columns for NA values

```{r}
raw_data %>%
     filter(if_any(everything(), is.na)) 
```

The NA value will be addressed later on

## Time to clean raw_data

### clean_names()

-   gives column 1 and 2 a label

-   puts an underscore "\_" in place of the space in Column 3

-   makes all uppercase letters into lower case

```{r}
clean_data <- raw_data %>% #add pipe to connect functions for one cohesive code chunk
  clean_names()
```

### rename()

-   Change the new name for Column 2 "x2" to "gene_id"

```{r}
clean_data <- raw_data %>%
  clean_names() %>% 
  rename(gene_id = x2)
```

### mutate() number 1

-   Here we change any NAs under the column *gsm3119015* to a 0

-   **NOTE: this is not the recommended approach for NAs in your dataset!**

    -   depends on your data and analysis but sometimes rows are removed or missing data are imputations are calculated

```{r}
clean_data <- raw_data %>%
  clean_names() %>% 
  rename(gene_id = x2) %>% 
  mutate(gsm3119015 = replace_na(gsm3119015, 0))
  #drop_na() %>% #use this to drop all rows with at least one NA
```

### mutate() number 2

-   Creates a new variable that combines the gene_id and gene_symbol together and separates them with "\|"; you will notice the new column is placed at the very end of the data frame

-   Notice how we use paste() from base R - meaning that you can use base R functions within the Tidyverse programming

```{r}
clean_data <- raw_data %>%
  clean_names() %>% 
  rename(gene_id = x2) %>% 
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  #drop_na() %>% #use this to drop all rows with at least one NA
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|"))
```

### select()

-   Makes gene_id_symbol the first column, keeps everything else in the same order using everything(), and deletes column 'x1' using the minus (-) operator

```{r}
clean_data <- raw_data %>%
  clean_names() %>% 
  rename(gene_id = x2) %>% 
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  #drop_na() %>% #use this to drop all rows with at least one NA
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|")) %>%
  select(gene_id_symbol, everything(), -x1, -gene_symbol, -gene_id)
```

## EXTRA processing steps to *harmonize* the study dataset with the metadata dataset

### Extra 1: make the first column the row IDs

-   Use column_to_rownames (from tibble package in the tidyverse) to change a column (this code is the first column) from a calculated variable to the row IDs

```{r}

clean_data <- raw_data %>%
  clean_names() %>% 
  rename(gene_id = x2) %>% 
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  #drop_na() %>% #use this to drop all rows with at least one NA
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|")) %>%
  select(gene_id_symbol, everything(), -x1, -gene_symbol, -gene_id) %>%
  column_to_rownames(var = colnames(.)[1])
```

### Extra 2: Change the order of the columns in *clean_data* based on the order of the values under *clean_meta_data*'s *sample_id* variable

```{r}
clean_data <- raw_data %>%
  clean_names() %>% 
  rename(gene_id = x2) %>% 
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  #drop_na() %>% #use this to drop all rows with at least one NA
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|")) %>%
  select(gene_id_symbol, everything(), -x1, -gene_symbol, -gene_id) %>%
  column_to_rownames(var = colnames(.)[1]) %>% 
  select(all_of(clean_meta_data$sample_id))
```

### Extra 3: run a quick check to see if the *clean_data* column names and the s*ample_id* values in  *clean_meta_data* match exactly

```{r}
setequal(clean_meta_data$sample_id, colnames(clean_data))
```

###  Save *clean_data_label, clean_data*, and *clean_meta_data* locally using here()

-   Write each to a CSV file

-   "processed_data" is the folder we are saving to

-   use the function read_csv to save "clean_data.csv" and "clean_meta_data.csv" clean CSV files - THIS WILL NOT INCLUDE THE ROW LABELS WE MADE

-   Use the function read.csv to save "clean_data_label.csv" to save a csv file that include our row labels

```{r}

write.csv(clean_data, here("processed_data", "clean_data_label.csv"))

write_csv(clean_data, here("processed_data", "clean_data.csv"))

write_csv(clean_meta_data, here("processed_data", "clean_meta_data.csv"))
```
