---
title: "Part 2: Working Locally with git & RStudio"
author: "Workshop Instructor"
format: html
editor: visual
---

# git in RStudio

## Introduction

In this section, we will get hands-on with **git** right inside RStudio. We'll learn the fundamental workflow of saving changes to our project's history. This entire process happens **locally**, on our own computers. We will explore this workflow in two ways: using RStudio's user-friendly graphical interface and using the powerful command line terminal.

**What we'll cover:**

1.  **Local git workflow** - Making and saving changes to your project history
2.  **Best practices** - Managing data files with `.gitignore` to keep repositories lean
3.  **Connecting to GitHub** - Publishing your local repository to GitHub.com as a remote backup and optionally sharing it publicly
4.  **Collaboration** - If time permits, we'll explore how to collaborate with colleagues using Pull Requests (also called Merge Requests), which allow you to propose changes to each other's projects safely and professionally

By the end of this section, you'll understand how to use git locally and how to leverage GitHub as both a backup and collaboration platform for your research projects.

## 1. Recap: The Initialized Repo

When your RStudio Project was created, the option to "Create a git repository" was selected. This means git has already been initialized and is tracking your project folder.

Your mission control for git in RStudio is the **git tab**, typically located in the top-right pane next to "Environment" and "History".

**Action:** Click on the **git tab**. You should see that the pane is currently empty, which means git doesn't see any unsaved changes.

### Understanding the git Tab Status Indicators

When you start making changes to files in your project, they'll appear in the git tab with colored status indicators. The git tab has **two columns** that show different information:

#### Left Column: Staging Area (What Will Be Committed)

| Color & Letter   | Meaning  | What It Tells You              |
|------------------|----------|--------------------------------|
| 🟢 **Green "A"** | Added    | New file staged for commit     |
| 🔵 **Blue "M"**  | Modified | Changed file staged for commit |
| 🔴 **Red "D"**   | Deleted  | File staged for deletion       |

#### Right Column: Working Directory (Current File Status)

| Color & Letter    | Meaning   | What It Tells You               |
|-------------------|-----------|---------------------------------|
| 🟡 **Yellow "M"** | Modified  | File has unstaged changes       |
| 🟡 **Yellow "?"** | Untracked | New file git doesn't know about |
| 🔴 **Red "D"**    | Deleted   | File deleted but not staged     |

#### Common Status Combinations You'll See

| Left Column | Right Column | What This Means                            |
|-------------|--------------|--------------------------------------------|
| *(empty)*   | Yellow "?"   | Brand new file, not tracked yet            |
| *(empty)*   | Yellow "M"   | Existing file changed, not staged          |
| Green "A"   | *(empty)*    | New file staged, ready to commit           |
| Blue "M"    | *(empty)*    | Modified file staged, ready to commit      |
| Blue "M"    | Yellow "M"   | Some changes staged, more changes unstaged |

::: callout-tip
## Quick Reference

-   **Left Column (Status)** = What's packed and ready for your commit
-   **Right Column** = What's changed but not packed yet
-   **Checking "Staged"** = Moves items from right column → left column
:::

## 2. The git Workflow in RStudio

The core git workflow involves three steps:

1.  You make changes to your files.
2.  You stage the specific changes you want to save.
3.  You commit those staged changes, creating a permanent snapshot in your project's history.

Let's walk through this by first saving the work you've already done in Parts 1 and 2 of the workshop.

### Confirm You Have a git Repository

Before we start committing files, let's make sure your project is actually set up with git and orient ourselves.

1.  **Check the git tab.** Look at the top-right pane of RStudio. Do you see a tab labeled "git" next to "Environment" and "History"? If yes, great! Your repository is initialized. If not, ask your instructor for help.

2.  **Open the Terminal.** Click on the **Terminal** tab (next to the Console tab at the bottom).

3.  **Verify the repository.** In the terminal, type:

``` bash
git status
```

You should see a message listing "Untracked files" - these are the R scripts and data files you've been working with throughout the workshop. This confirms git is watching your project folder but hasn't saved any snapshots yet.

4.  **Check what files you have.** Let's see what we're working with. In the terminal, type:

``` bash
ls
```

You should see your workshop files like `ARCS_Part1.R`, `ARCS_Part2.R`, folders like `raw_data/`, `scripts/`, etc.

**Perfect! You have a git repository and files ready to commit. Let's save your first snapshot.**

### Your First Commit: Saving Existing Work

Before creating new files, let's commit the scripts and data files you've been working with throughout the workshop series.

1.  **Check the git tab.** Click on the **git tab** in the top-right pane. You should see several files listed with yellow "?" icons, indicating they are untracked (new files that git doesn't know about yet). You might see files like:
    -   `ARCS_Part1.R` and `ARCS_Part1b.R`
    -   `ARCS_Part2.R` and `ARCS_Part2b.R`
    -   `Shank3_metadata_parsed.csv`
    -   Various files in your `raw_data/` folder
2.  **Stage all files.** Instead of clicking each file individually, let's use the command line to add everything at once. Open the **Terminal** tab (next to the Console) and type:

``` bash
git add .
```

The `.` means "add all changed files in the current directory and sub directories."

::: callout-note
## About Line Ending Warnings

You may see warnings like "CRLF will be replaced by LF the next time Git touches it." This is normal! Git is automatically converting Windows-style line endings to Unix-style for consistency across different operating systems. You can safely ignore these warnings.
:::

3.  **Check the git tab again.** Look back at the git tab. All your files should now show green "A" icons in the left column, meaning they're staged and ready to commit.

4.  **Commit your work.** Click the **"Commit"** button at the top of the git pane.

5.  **Write a clear commit message.** In the commit window that opens, type this message in the "Commit message" box:

    `Add workshop scripts and data from Parts 1 and 2`

6.  **Complete the commit.** Click the **"Commit"** button. A dialog will confirm your changes. Close it when done.

7.  **View the history.** Click the **"History"** button (looks like a clock) at the top of the git pane. You can see your first commit with all the files you've been working on!

**Congratulations, you've created your first snapshot of the project!**

### Creating New Files for Practice

Now let's create a new analysis script for the examples in this section.

1.  **Create a new file.** In RStudio, go to File → New File → R Script and save it as `ARCS_Part3b.R` in your `scripts` folder.

2.  **Add a comment.** Add this line to the script:

```{r}
#| eval: false
# This is a placeholder for our script content
```

3.  **View the change in the git tab.** Click back to the **git tab**. You'll see `ARCS_Part3b.R` listed with a yellow "?" icon (untracked file).

4.  **Stage the file.** Click the checkbox under the "Staged" column for `ARCS_Part3b.R`. The icon will change to a green "A".

5.  **Commit the change.** Click **"Commit"**, write the message `Add initial analysis script`, and click **"Commit"** again.

**You've now completed your second version control cycle!**

## 3. Excluding Data Files: The .gitignore File

Before we continue, let's discuss an important best practice: **not storing large data files in git repositories**.

### Why Exclude Data Files?

Git is designed to track *code* and *text files*, not large data files. Here's why:

-   **Repository bloat:** Data files can make your repository huge and slow to clone/push/pull
-   **Version control inefficiency:** Git tracks every version of every file. A 100MB dataset becomes 500MB after 5 versions!
-   **GitHub limitations:** GitHub has file size limits (100MB per file) and repository size recommendations (\< 1GB)
-   **Not designed for data:** Git compresses text efficiently but struggles with binary data formats

::: callout-note
## What About Data Storage?

For data files, consider these alternatives:

-   **Small reference data (\< 10MB):** Can include in repository if rarely changes
-   **Medium data (10MB - 1GB):** Use cloud storage (Dropbox, Google Drive, OneDrive) and document the location
-   **Large data (\> 1GB):** Use institutional data repositories, OSF, Zenodo, or Figshare
-   **Very large data:** Consider **Git LFS (Large File Storage)** - a Git extension that stores large files externally while keeping references in Git. See [git-lfs.github.com](https://git-lfs.github.com)
:::

### Creating a .gitignore File

The `.gitignore` file tells git which files and folders to ignore completely.

1.  **Create the file.** In RStudio, go to File → New File → Text File

2.  **Add these lines:**

``` gitignore
# Ignore data folders
raw_data/
processed_data/

# Ignore large output files
outputs/*.pdf
outputs/*.html

# Keep the folder structure but ignore contents
!raw_data/.gitkeep
!processed_data/.gitkeep
!outputs/.gitkeep
```

3.  **Save the file** as `.gitignore` (with the dot!) in your project root folder

4.  **Create placeholder files** so the folders still exist in git. In the terminal:

``` bash
touch raw_data/.gitkeep
touch processed_data/.gitkeep
touch outputs/.gitkeep
```

The `.gitkeep` files are empty placeholders that ensure the folder structure is preserved in your repository, even though the data files are ignored.

5.  **Commit the .gitignore file.** In the git tab, you'll see `.gitignore` and the three `.gitkeep` files as new files. Stage and commit them with the message: `Add .gitignore to exclude data files`

::: callout-tip
## Check What's Being Tracked

After adding `.gitignore`, run this in the terminal to verify your data files are ignored:

``` bash
git status --ignored
```

You should see your data folders listed under "Ignored files".
:::

**Key Takeaway:** Keep your git repository lean by tracking code, not data. Document where your data lives in your README file instead!

## 4. git on the Command Line (CLI) - A Quick Look

The RStudio git tab is a fantastic visual tool, but it's just running git commands for you in the background. Let's pull back the curtain and run those same commands ourselves.

1.  **Open the Terminal.** In RStudio, there is a tab next to the "Console" tab labeled "Terminal". Click on it.

2.  **Check the project status.** The single most useful git command is `git status`. It tells you the current state of your repository.

    **Action:** In the terminal, type `git status` and press Enter.

    **Result:** It should say `nothing to commit, working tree clean`. This is because we just saved all our changes with the last commit.

3.  **Make another change.** Go back to your `ARCS_Part3b.R` script and add a new line of code.

```{r}
#| eval: false
# This is our first tracked change.
print("Hello from R!")
```

4.  **Check the status again.** Go back to the terminal and run `git status` again. Notice that git now sees that `ARCS_Part3b.R` has been modified.

5.  **Stage the change (CLI).** To stage the file using the command line, we use `git add`.

    **Action:** In the terminal, type `git add .` and press Enter. The `.` is shorthand for "all changed files in the current directory."

    **Check:** Run `git status` one more time. You'll see the file is now listed under "Changes to be committed."

6.  **Commit the change (CLI).** To commit, we use `git commit` with the `-m` flag to provide a message.

    **Action:** In the terminal, type `git commit -m "Add print statement via CLI"` and press Enter.

**You've just done the exact same workflow as before, but using the command line!**

**Key Takeaway:** The RStudio git tab is a user-friendly interface for running these fundamental git commands. You can use whichever you prefer, but it's good to know what's happening behind the scenes.

## 5. Practice: Making Commits with GUI and CLI

Now let's create a complete analysis script by making two commits: one using the RStudio GUI and one using the command line. Each commit will add a meaningful piece of functionality to your script.

### Commit 1: Load and Explore Data (GUI Method)

Let's add code to load our cleaned datasets and explore their structure.

1.  **Add data loading code.** Open your `ARCS_Part3b.R` file and add these lines:

```{r}
#| eval: false
# ==============================================================================
# ARCS Workshop Part 3 - Gene Expression Analysis
# This script demonstrates data loading and visualization with git version control
# ==============================================================================

# Load the tidyverse package (includes readr, dplyr, ggplot2, tidyr, stringr)
library(tidyverse)

# ------------------------------------------------------------------------------
# 1. LOAD DATA
# ------------------------------------------------------------------------------

# Load cleaned gene expression data from the processed_data folder
# read_csv() is from the readr package (part of tidyverse)
# It's faster and has better defaults than base R's read.csv()
gene_data <- read_csv("processed_data/clean_data_label.csv")

# Load the sample metadata (information about each sample)
# This includes genotype, experimental condition, etc.
metadata <- read_csv("processed_data/clean_meta_data.csv")

# ------------------------------------------------------------------------------
# 2. EXPLORE DATA STRUCTURE
# ------------------------------------------------------------------------------

# Display basic information about the dataset
# The first column contains gene identifiers, remaining columns are sample expression values
cat("Dataset dimensions:", nrow(gene_data), "genes x", (ncol(gene_data) - 1), "samples\n")

# Display the first few rows and first 5 sample columns
# select() is from dplyr (part of tidyverse) and lets us choose specific columns
# 1:6 selects the gene name column plus the first 5 sample columns
gene_data %>%
  select(1:6) %>%
  head()

# Check metadata structure
cat("\nMetadata structure:\n")
glimpse(metadata)

# ------------------------------------------------------------------------------
# 3. DATA ANALYSIS AND VISUALIZATION
# (Additional analysis code will be added here)
# ------------------------------------------------------------------------------
```

2.  **Check the git tab.** You'll see `ARCS_Part3b.R` appears with a blue "M" (modified).

3.  **Review your changes (Optional but helpful!).** Click on the filename `ARCS_Part3b.R` in the git tab. A "diff" window will open showing your new code in green. This is incredibly useful for reviewing your work before committing.

4.  **Stage and commit.** Check the box to stage the file, click "Commit", and write the message: `Add data loading and exploration`

::: callout-tip
## GUI Tip: Reviewing Changes Before Committing

Always review the diff before committing! The green highlighting shows what you're adding, and you can spot any mistakes, debug code, or unwanted changes before they become part of your project history.
:::

### Commit 2: Add Visualization (CLI Method)

Now let's add a publication-quality visualization to complete our analysis script.

1.  **Add plotting code.** Add this to your `ARCS_Part3b.R` file, replacing the placeholder comment in Section 3:

```{r}
#| eval: false
# ------------------------------------------------------------------------------
# 3. DATA ANALYSIS AND VISUALIZATION
# ------------------------------------------------------------------------------

# Filter for a specific gene: Gnai3 (a G-protein signaling gene)
# We use filter() and str_detect() to find the gene by name, which is more robust
# than using slice(1) because it works regardless of data sorting
gnai3_data <- gene_data %>% 
  filter(str_detect(names(gene_data)[1], "Gnai3") | 
         str_detect(.[[1]], "Gnai3"))  # Search in the first column for "Gnai3"

# Alternative approach if the first column has a name:
# gnai3_data <- gene_data %>% filter(str_detect(gene_id, "Gnai3"))

# Reshape the data for ggplot2 using tidyr's pivot_longer()
# This converts wide format (one column per sample) to long format (one row per sample)
# Long format is preferred for ggplot2 and most tidyverse functions
gnai3_long <- gnai3_data %>%
  pivot_longer(
    cols = -1,                    # Select all columns except the first (gene name)
    names_to = "sample_id",       # Create a new column called "sample_id" with column names
    values_to = "expression"      # Create a new column called "expression" with the values
  ) %>%
  # Join with metadata to get experimental conditions
  # This connects expression values with genotype and condition information
  left_join(metadata, by = "sample_id")

# Create a boxplot grouped by experimental condition
# This shows how Gnai3 expression varies across genotypes and conditions
gnai3_plot <- ggplot(gnai3_long, aes(x = geno_cond, y = expression, fill = genotype)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +      # Add boxplot layer (alpha = transparency)
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +    # Add individual data points
  scale_fill_manual(values = c("WT" = "lightblue", "Shank3" = "salmon")) +  # Custom colors
  labs(
    title = "Gnai3 Gene Expression by Genotype and Condition",
    x = "Experimental Group",
    y = "Read Counts",
    fill = "Genotype"
  ) +
  theme_minimal() +                                     # Clean, modern theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold title
    axis.text.x = element_text(angle = 45, hjust = 1)       # Rotate x-axis labels for readability
  )

# Display the plot
print(gnai3_plot)

# Summary statistics by group
gnai3_summary <- gnai3_long %>%
  group_by(geno_cond, genotype) %>%
  summarise(
    mean = mean(expression),
    sd = sd(expression),
    n = n(),
    .groups = "drop"
  )

cat("\nGnai3 Expression Summary Statistics:\n")
print(gnai3_summary)

# ==============================================================================
# END OF SCRIPT
# Ready for additional analysis and plotting
# ==============================================================================
```

2.  **Use the Terminal workflow:**

    -   Check status: `git status`
    -   Stage the file: `git add scripts/ARCS_Part3b.R`
    -   Check again: `git status`
    -   Commit: `git commit -m "Add Gnai3 expression visualization and summary statistics"`

3.  **Verify your work.** Click the "History" button in the git tab (or run `git log --oneline` in the terminal). You should now see all your commits listed!

::: callout-note
## When to Use GUI vs. CLI?

-   **GUI:** Great for visual learners, reviewing changes before committing, and when you're just starting out.
-   **CLI:** Faster once you learn it, more powerful for advanced operations, and works everywhere (even on remote servers).

Many experienced users mix both approaches: GUI for reviewing diffs, CLI for quick commits!
:::

## 6. Local-First vs. Remote-First Repositories

This is a perfect time to discuss the two main ways a git project begins.

### Local-First (What we just did):

-   **The process:** You start a project on your own computer first. You initialize a git repository, work on your files, and create a history of commits, all locally.
-   **The goal:** Later, you can decide to "push" this local project to a remote server like gitHub to back it up, share it, or collaborate.
-   **Analogy:** You write a new document in Microsoft Word on your desktop. When you're ready, you upload it to OneDrive to share with your colleagues.

### Remote-First (Cloning):

-   **The process:** The project already exists on a remote server like gitHub. Your first step is to "clone" it, which downloads a complete copy of the project and its entire version history onto your computer.
-   **The goal:** This is the standard way to join a project that a colleague has already started or to contribute to an open-source project.
-   **Analogy:** A colleague shares a document with you via OneDrive. You use the "Open in Desktop App" feature to download a local copy to your computer so you can start editing.

**In this workshop, we started "local-first." Later, in the collaboration section, you will experience the "remote-first" workflow when you clone your partner's repository.**

## 7. Publishing Your Repository to gitHub

Now that we have a local repository with a commit history, let's back it up and share it by publishing to gitHub.com. This requires two steps: setting up authentication with SSH keys, and creating a remote repository.

### Step 1: Set Up SSH Authentication

SSH (Secure Shell) is a secure way to connect your computer to gitHub. Think of it as a special key that proves your identity without typing your password every time.

::: callout-note
## Alternative: Personal Access Tokens (PATs)

Personal Access Tokens are an alternative authentication method using HTTPS instead of SSH. However, they expire every 90 days and require renewal, which can disrupt your workflow. **SSH is strongly preferred for long-term research projects** because it's a one-time setup.

If SSH doesn't work due to network/firewall restrictions, gitHub Desktop (which uses HTTPS) is a good backup option, or consult your instructor about PAT setup.
:::

#### Check for Existing SSH Keys

First, let's see if you already have SSH keys on your computer.

**In the RStudio Terminal**, run:

``` bash
ls ~/.ssh
```

If you see files named `id_ed25519` or `id_rsa`, you already have keys! Skip to "Add Your Key to gitHub" below. If not, continue to generate new keys.

#### Generate a New SSH Key

::::: panel-tabset
## macOS/Linux

1.  **Generate the key.** In the Terminal, run this command (replace with your actual gitHub email):

``` bash
ssh-keygen -t ed25519 -C "your_email@example.com"
```

2.  **Save the key.** When prompted "Enter a file in which to save the key," press **Enter** to accept the default location (`~/.ssh/id_ed25519`).

3.  **Set a passphrase (optional).** You'll be asked to enter a passphrase. This adds extra security but is optional. You can press **Enter** twice to skip it, or type a passphrase you'll remember.

4.  **Start the SSH agent:**

``` bash
eval "$(ssh-agent -s)"
```

You should see something like `Agent pid 12345`.

5.  **Add your key to the SSH agent:**

``` bash
ssh-add ~/.ssh/id_ed25519
```

6.  **Copy your public key to the clipboard:**

``` bash
pbcopy < ~/.ssh/id_ed25519.pub
```

Your key is now copied and ready to paste into gitHub!

## Windows

::: callout-note
## Windows Users: Use git Bash or RStudio Terminal

These commands work in **git Bash** (installed with git for Windows) or the **RStudio Terminal**. If you're using Command Prompt or PowerShell, some commands may differ. We recommend using the RStudio Terminal for consistency.
:::

1.  **Generate the key.** In the Terminal, run this command (replace with your actual gitHub email):

``` bash
ssh-keygen -t rsa -b 4096 -C "your_email@example.com"
```

2.  **Save the key.** When prompted "Enter a file in which to save the key," press **Enter** to accept the default location.

    You'll see: `Your identification has been saved in /c/Users/YourName/.ssh/id_rsa`

3.  **Set a passphrase (optional).** You'll be asked to enter a passphrase. This adds extra security but is optional.

    -   To skip: Press **Enter** twice
    -   To add security: Type a passphrase you'll remember, press **Enter**, then type it again

4.  **Start the SSH agent:**

``` bash
eval "$(ssh-agent -s)"
```

You should see something like `Agent pid 12345`.

5.  **Add your key to the SSH agent:**

``` bash
ssh-add ~/.ssh/id_rsa
```

You should see: `Identity added: /c/Users/YourName/.ssh/id_rsa`

6.  **Copy your public key to the clipboard:**

**Option 1 (git Bash/RStudio Terminal):**

``` bash
cat ~/.ssh/id_rsa.pub | clip
```

**Option 2 (If clip doesn't work):**

``` bash
cat ~/.ssh/id_rsa.pub
```

Then manually select and copy the entire output (it starts with `ssh-rsa` and ends with your email).

**Option 3 (PowerShell users):**

``` powershell
Get-Content ~/.ssh/id_rsa.pub | Set-Clipboard
```

Your key is now copied and ready to paste into gitHub!

::: callout-warning
## Windows Troubleshooting

If `ssh-agent` won't start, try opening RStudio or git Bash **as Administrator**. Right-click the application icon and select "Run as administrator."

If you see "Could not open a connection to your authentication agent," run this first:

``` bash
ssh-agent bash
```

Then retry the `ssh-add` command.
:::
:::::

::: callout-important
## Keep Your Keys Safe!

-   The `.pub` file is your **public key** (safe to share with gitHub)
-   The file without `.pub` is your **private key** (never share this!)
-   If someone asks for your SSH key, they always mean the public one
:::

#### Add Your SSH Key to gitHub

Now we'll add your public key (which is copied to your clipboard) to your gitHub account:

1.  Go to [github.com](https://github.com) and log in to your account

2.  Click on your **profile photo** in the upper-right corner

3.  Click **Settings**

4.  In the left sidebar, click **SSH and GPG keys**

5.  Click the green **New SSH key** button

6.  In the **Title** field, add a descriptive name like "My RStudio Laptop"

7.  In the **Key** field, paste your key (Cmd+V on Mac, Ctrl+V on Windows)

8.  Click **Add SSH key**

9.  If prompted, confirm your gitHub password

#### Test Your Connection

Back in the RStudio Terminal, run:

``` bash
ssh -T git@github.com
```

You may see a warning about authenticity - type `yes` and press Enter.

You should see a message like: `Hi username! You've successfully authenticated...`

**Success!** Your computer can now securely talk to gitHub.

### Step 2: Create a Repository on gitHub.com

Now let's create a home for your local repository on gitHub:

1.  Go to [github.com](https://github.com) and log in

2.  Click on your **profile picture** in the top-right corner

3.  Select **Your repositories**

4.  Click the green **New** button

5.  **Repository name:** Enter a name (e.g., `ARCS-workshop`)

6.  **Description (optional):** Add a brief description like "Reproducibility workshop project"

7.  **Visibility:** Choose **Private** (or Public if you want others to see it)

8.  **Important:** Do NOT check "Add a README file" - we already have files locally!

9.  Click the green **Create repository** button

### Step 3: Connect Your Local Repository to gitHub

gitHub will show you a page with setup instructions. We'll use the commands for "push an existing repository from the command line."

In your RStudio Terminal, run these commands one at a time:

1.  **Add the remote connection** (replace `username` and `yourreponame` with yours):

``` bash
git remote add origin git@github.com:username/yourreponame.git
```

This tells git where the remote repository lives.

2.  **Ensure you're on the main branch:**

``` bash
git branch -M main
```

3.  **Push your commits to gitHub:**

``` bash
git push -u origin main
```

The `-u` flag sets up tracking so future pushes are easier.

You should see output showing your commits being uploaded. When it finishes, refresh your gitHub repository page in your browser - you should see all your files and commit history!

::: callout-tip
## Troubleshooting: Using gitHub Desktop as a Backup Method

If you encounter authorization issues with the command line, you can use gitHub Desktop as an alternative:

1.  Download and install [gitHub Desktop](https://desktop.github.com/)
2.  Sign in with your gitHub account
3.  Click "Add" → "Add Existing Repository"
4.  Navigate to your project folder
5.  Click "Publish repository" to upload to gitHub.com

This method uses HTTPS authentication instead of SSH.
:::

**Congratulations!** Your local repository is now backed up on gitHub. Any changes you commit locally can be pushed to gitHub with `git push`, and you can access your code from anywhere.

### GitHub Enterprise Options at Yale

If you're working with sensitive research data or need enhanced security and compliance features, Yale offers two GitHub Enterprise options beyond the standard GitHub.com platform:

#### GitHub Enterprise Server (On-Premises)

GitHub Enterprise Server is a **self-hosted platform** for managing research code and data workflows within Yale's secure infrastructure. Your research team can use GitHub Enterprise Server to version control analysis scripts, collaborate on computational methods, maintain reproducible research pipelines, and integrate with Yale's research computing resources—all while maintaining strict data security and compliance standards.

**Key Features:**

\- Hosted on Yale's own infrastructure

\- Developers familiar with GitHub.com can onboard and contribute seamlessly using familiar features and workflows

\- **Data Classification Level Supported: High** (suitable for sensitive research data)

\- Full control over data location and security policies

\- Ideal for projects requiring institutional data governance

**When to use:** Projects involving sensitive health data, HIPAA-compliant research, or any work requiring High-level data classification.

#### GitHub Enterprise Cloud

GitHub Enterprise Cloud adds **advanced functionality to GitHub.com**, such as:

\- SAML authentication for single sign-on with Yale credentials

\- Additional GitHub Actions minutes for automated workflows

\- Ability to restrict email notifications to verified domains

\- Privately published GitHub Pages sites for internal documentation

\- Enhanced security and compliance features

**Key Features:**

\- Cloud-based (managed by GitHub, not self-hosted)

\- Seamless integration with Yale's authentication systems

\- Advanced collaboration and security tools

\- No infrastructure management required

\- **Data Classification Level Supported: Moderate** (not suitable for high-sensitivity data)

**When to use:** Collaborative research projects that need enhanced security features but work with public or moderate-sensitivity data.

::: callout-tip
## Choosing the Right GitHub Platform

-   **GitHub.com (Free/Pro):** Public research, open-source projects, teaching repositories
-   **GitHub Enterprise Cloud:** Collaborative research with Yale authentication, moderate-sensitivity data
-   **GitHub Enterprise Server (On-Premises):** High-sensitivity data, HIPAA compliance, institutional data governance requirements

Contact [Yale ITS](https://its.yale.edu/) for access to GitHub Enterprise options and guidance on which platform fits your research needs.

**GitHub Enterprise Service & Support:** [Request access or support via Yale ServiceNo](https://yale.service-now.com/it?id=service_offering&sys_id=bc29b84997f88a506f0430671153af28)
:::

## 8. *Bonus:* Collaboration with gitHub Desktop & RStudio

Now that we've learned the basics of git and gitHub, let's explore one of the most powerful aspects of version control: **collaboration**. In this section, we'll demonstrate a complete, real-world collaborative workflow where one person proposes a change to another person's project.

This workflow is essential for:

\- Contributing to open-source projects

\- Collaborating with colleagues on research code

\- Proposing improvements to shared analysis scripts

\- Working on team projects with proper review processes

### Understanding the Collaboration Roles

In any collaborative git workflow, there are typically two key roles:

-   **Project Owner:** The person who owns the original repository and has the authority to accept or reject changes
-   **Contributor:** The person who wants to suggest improvements, fixes, or additions to the project

### The Fork-and-Pull Request Workflow

Unlike giving someone direct write access to your repository (which can be risky), gitHub's **fork-and-pull request** workflow provides a safe way to collaborate:

1.  **Fork:** The contributor creates their own complete copy of the project
2.  **Clone:** They download their copy to work on locally
3.  **Branch:** They create a separate branch for their specific contribution
4.  **Edit:** They make their changes and commit them
5.  **Push:** They upload their changes to their fork on gitHub
6.  **Pull Request:** They formally ask the project owner to review and merge their changes

Let's walk through this process step by step.

### Part 1: The Contributor's Workflow

This demonstrates how to contribute to a project you don't have direct write access to.

#### Step 1: Fork the Repository

Forking creates a personal copy of someone else's project under your own gitHub account.

**Actions:**

1\. Navigate to the project owner's repository on gitHub.com

2\. Click the **Fork** button in the top-right corner of the repository page

3\. gitHub will create a new repository under your account with the same name

4\. You now have your own complete copy that you can freely modify

::: callout-note
## What Forking Actually Does

When you fork a repository, gitHub creates an exact copy of the entire project history under your account. This includes all commits, branches, and files. Think of it as creating your own workspace where you can experiment freely without affecting the original project.
:::

#### Step 2: Clone Your Fork to gitHub Desktop

Now you need to get the project files onto your local computer to work with them.

**Actions:**

1\. In your newly forked repository, click the green **\<\> Code** button

2\. Copy the HTTPS URL that appears

3\. Open **gitHub Desktop**

4\. Go to **File \> Clone Repository** 5. Paste the URL in the URL tab

6\. Choose where you want the project saved on your computer

7\. Click **Clone**

The entire project is now downloaded to your computer and ready for editing.

#### Step 3: Create a New Branch

Creating a separate branch for your contribution is a best practice that keeps your changes isolated and organized.

**Actions:**

1\. In gitHub Desktop, with your cloned repository selected, click on the **Current Branch** dropdown (it should say "main")

2\. Click the **New Branch** button

3\. Give your branch a descriptive name that explains what you're adding (e.g., `add-summary-plot`, `fix-data-loading`, `improve-documentation`)

4\. Click **Create Branch**

5\. Click **Publish branch** to upload this new, empty branch to your fork on gitHub

::: callout-tip
## Branch Naming Conventions

Good branch names are:

\- **Descriptive:** `add-ggplot-visualization` instead of `new-feature`

\- **Verb-focused:** Start with action words like `add`, `fix`, `update`, `remove`

\- **Lowercase with hyphens:** `improve-error-handling` instead of `Improve_Error_Handling`
:::

#### Step 4: Make Your Edits in RStudio

Time to contribute your improvement to the project!

**Actions:**

1\. In gitHub Desktop, click the **Open in RStudio** button (or navigate to the project folder and open the .Rproj file)

2\. Create a new file to add documentation to the project. Go to **File → New File → Markdown File**

3\. Save the file as `README.md` in the project root folder

4\. Add the following content to document the workshop project:

``` markdown
# Introduction to Programming in R: A Workshop Series 💻

Welcome! This repository was created as part of the 3-part "Introduction to Programming in R" workshop series, part of the **Advancing Reproducibility with Computational Skills (ARCS)** initiative at the Yale Medical Library.

This series is designed to guide you from the fundamentals of R programming to more advanced topics, including data manipulation with the Tidyverse and version control with Git/GitHub for collaborative and reproducible research.

------------------------------------------------------------------------

## Part 1: Fundamentals and Project Management

### Summary

If you are new to coding or have never used R before, this workshop is for you! In this introduction to R fundamentals, you will complete hands-on coding exercises, follow along with more complex R scripts, and connect these skills to reproducible research practices.

### In this workshop, you will learn to:

-   Navigate the RStudio interface and execute basic R commands
-   Apply fundamental programming concepts (objects, functions, basic operations)
-   Create vectors, data frames, and matrices
-   Find a dataset in a public repository and read dataset files into RStudio
-   Implement file organization and naming conventions for reproducible research
-   Understand the advantages of R for research workflow automation

### Prerequisites:

-   A computer where you have the necessary permissions to install software
-   [R](https://cran.r-project.org/) and [RStudio](https://posit.co/download/rstudio-desktop/) installation

------------------------------------------------------------------------

## Part 2: Data Structures and Reports

### Summary

If you want to start learning how to use R to analyze data and be able to report your findings, this workshop is for you! Using the same dataset from Part 1, you will learn how R stores data, how to make choices as you set up your data for analysis, and how to create a reproducible record of your work.

### In this workshop, you will learn to:

-   Manipulate vectors, lists, matrices, and data frames
-   Identify the advantages and limitations of different data structures in R
-   Implement literate programming principles in data analysis
-   Generate dynamic reports using Quarto for reproducible documentation

### Prerequisites:

-   Completion of Part 1: Fundamentals and Project Management
-   [Quarto](https://quarto.org/docs/get-started/) installation

------------------------------------------------------------------------

## Part 3: Tidyverse Packages and Git

### Summary

If you want to start leveraging the R packages ecosystem while keeping track of how you do it, this workshop is for you! Using the same dataset from Parts 1 & 2, you will learn how to transform and "tidy" your data using the Tidyverse, and extend your code management skills by learning to collaborate effectively using Git for version control.

### In this workshop, you will learn to:

-   Understand how to subset data effectively
-   Execute complex data transformations using `dplyr` (`filter`, `select`, `mutate`, `summarize`, `group_by`) and other useful packages
-   Initialize Git repositories and commit project changes
-   Understand version control and collaboration methods using GitHub

### Prerequisites:

-   Completion of Part 2: Data Structures and Reports
-   [GitHub Account](https://github.com/signup)
-   [GitHub Desktop](https://desktop.github.com/)
-   Windows users only: [Git Bash](https://git-scm.com/downloads/)
-   **Note:** Please accept all defaults for installer prompts.
```

5.  **Save your changes** (Ctrl+S / Cmd+S)

::: callout-note
## Why Add a README?

A README file is the first thing people see when they visit a repository on GitHub. It should explain: - What the project is about - How to get started - What prerequisites are needed - How the project is organized

Good documentation makes your research more accessible and reproducible!
:::

#### Step 5: Commit and Push Your Changes

Now save your work to the project's version history.

**Actions:**

1\. Switch back to **gitHub Desktop**

2\. You should see your modified files listed in the left panel

3\. Review the changes by clicking on the file names to see the exact differences

4\. In the **Summary** box at the bottom-left, write a clear commit message describing your contribution (e.g., "Add README documentation for workshop series")

5\. Optionally, add a longer description in the **Description** box

6\. Click **Commit to \[your-branch-name\]**

7\. Click the **Push origin** button to upload your committed changes to your fork on gitHub

#### Step 6: Open a Pull Request

A Pull Request (PR) is your formal way of asking the project owner to review and potentially merge your contribution into their project.

**Actions:**

1\. After pushing, a blue banner should appear in gitHub Desktop saying "Create Pull Request"

2\. Click **Create Pull Request** - this opens a page on gitHub.com

3\. gitHub automatically sets up the request to merge your branch into the original project's main branch

4\. Give your Pull Request a clear, descriptive title

5\. In the description box, explain: - What changes you made - Why you made them - Any relevant details the reviewer should know

6\. Click **Create pull request**

**Example PR description:**

```         
## Summary
Added comprehensive README documentation to help users understand the ARCS workshop series structure, learning objectives, and prerequisites for each part.

## Changes Made
- Created README.md file in the project root
- Documented all three parts of the workshop series (Fundamentals, Data Structures, and Tidyverse/Git)
- Listed clear learning objectives for each workshop
- Included prerequisite requirements and installation links
- Added professional formatting with headers and sections

## Why This Improvement
The repository lacked documentation explaining:
- What the workshop series covers
- How the three parts build on each other
- What software and accounts participants need
- The connection to reproducible research practices

This README makes the project more accessible to new learners and provides context for anyone discovering this repository on GitHub.
```

### Part 2: The Project Owner's Review Process

This demonstrates how a project maintainer handles incoming contributions.

#### Step 7: Review and Manage the Pull Request

The project owner receives notification of the Pull Request and can now review the proposed changes.

**Actions for the Project Owner:**

1\. Navigate to their original repository on gitHub.com

2\. Click on the **Pull requests** tab to see all incoming requests

3\. Click on the specific PR to review it in detail

4\. In the **Files changed** tab, review the exact code changes:

\- Green lines show additions

\- Red lines show deletions

\- Can leave line-by-line comments for discussion

5\. In the **Conversation** tab, can:

\- Ask questions about the implementation

\- Request changes or improvements

\- Approve the changes

\- Discuss the contribution with the contributor

**Review Options:**

\- **Approve and Merge:** If the changes look good, click **Merge pull request** then **Confirm merge**

\- **Request Changes:** Leave comments asking for modifications before merging

\- **Close without Merging:** If the contribution doesn't fit the project's direction

::: callout-important
## Code Review Best Practices

When reviewing Pull Requests:

\- **Be constructive:** Focus on the code, not the person

\- **Be specific:** Point out exact lines and suggest improvements

\- **Be appreciative:** Thank contributors for their time and effort

\- **Test locally:** For important changes, pull the branch and test it yourself
:::

#### After Merging: Cleanup and Next Steps

Once a Pull Request is merged:

**For the Project Owner:**

\- The contribution is now part of the main project

\- The contributor's branch can be safely deleted

\- Consider adding the contributor to the project's acknowledgments

**For the Contributor:**

\- Switch back to the main branch and pull the latest changes to get your merged contribution

\- Delete your feature branch (it's no longer needed)

\- Your contribution is now part of the official project history!

### Key Benefits of This Workflow

This fork-and-pull request workflow provides several advantages:

-   **Safety:** The original project is never at risk during development

-   **Review:** All changes go through a review process before integration

-   **Discussion:** Contributors and maintainers can discuss improvements openly

-   **History:** Every contribution is documented with clear attribution

-   **Learning:** Contributors learn from feedback and improve their skills

### When to Use This Workflow

This collaboration pattern is ideal for:

\- **Open-source contributions:** Adding features to public R packages or analysis scripts

\- **Research collaboration:** Multiple researchers contributing to shared analysis code

\- **Code review:** Ensuring quality and catching errors before integration

\- **Knowledge sharing:** Learning from others' coding approaches and receiving feedback

**Congratulations!** You now understand the complete collaborative git workflow that powers most modern software development and is increasingly important in reproducible research.
