---
title: "Tidyverse"
sidebar: arcs03
filters:
  - webr
webr:
  cell-options:
    reset-button: true
---

## Overview

In this session we will cover data importing and exporting using the `rio` package along with cleaning and validation with Tidyverse.

::: callout-note
## By the end of this section, you should be able to...

- Use the `here` package to build file paths relative to your project root
- Use rio::import() and rio::export() to streamline reading and writing diverse file formats
- Explain the difference between `read_csv()` and `read.csv()`, and what a tibble is
- Inspect a data frame using `view()`, `head()`, and `str()`
- Use the pipe operator `%>%` to chain multiple data cleaning steps together
- Clean column names automatically using `clean_names()`
- Rename, recode, and create new columns using `rename()`, `mutate()`, and `case_when()`
- Combine two columns into a new variable using `paste0()`
- Select and reorder columns using `select()` and `everything()`
- Handle missing values using `replace_na()`
- Harmonize two data frames so their columns and row identifiers align
:::

------------------------------------------------------------------------

## What are R Packages

- Packages are collections of functions and data that extend the functionality of the base R programming language.

- They are used to perform specialized tasks, from creating graphs to running complex statistical models.

- Packages are a core reason for R's popularity, as they are created and shared by a global community of users.

## Installing packages

You can easily install (most of) packages using the install.packages() function directly in R. For example:

One at a time:

```{r}
install.packages("tidyverse")
```

Multiple at once:

```{r}
install.packages(c("tidyverse", "magrittr", "here", "readr", "janitor"))
```

## Load Packages

After installing any package or after you restart R, you need to load them into your R session using the `library()` function. For example:

One at a time:

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(readr)
library(janitor)
library(rio)
```

------------------------------------------------------------------------

## Importing data into R using Rio() and Here()

## Quick background

### Base R (no packages)

- This is the original, old school way of reading `.csv`, `.txt`, and `.dat` files and returns a data frame in the R environment using the function read.*csv,* read.*table*, etc.

- Cumbersome - it requires additional manual coding specifications based on the file type; it can be slow for large datasets; and can lead to unexpected behaviors with certain column types (i.e. converting character columns into factors)

### readr package

- Modern alternative to *Base R* and requires less configuration while being faster, cleaner, and provides less coding headaches (its does not convert character columns into factors) and uses the function read\_*csv,* read\_*tsv,* etc.

- Returns a *tibble,* a modern data frame with features that make it print nicely, displays on a screen better, does not change variables names, and does not use row labels. The output when inspecting the data set using *str()* are nicer looking and more informative

::: callout-note
## `read_csv()` vs `read.csv()`

This section uses `read_csv()` from the Tidyverse rather than `read.csv()` from base R. Key differences:

| Feature      | `read_csv()`                  | `read.csv()`   |
|--------------|-------------------------------|----------------|
| Column types | Explicit (`col_double`, etc.) | Inferred       |
| Speed        | Faster on large files         | Slower         |
| Output       | **Tibble**                    | Data frame     |
| Column names | Preserved as-is               | May be altered |
| Row labels   | Not used                      | Supported      |

A **tibble** is a modern data frame that prints more cleanly, displays better on screen, and produces more informative output from functions like `str()`.
:::

### haven package

- The **`haven`** package is the dedicated tool for reading proprietary formats like SPSS (`.sav`), SAS (`.xpt` & `.sas7bdat`), and Stata (`.dta`), and is a key part of the Tidyverse ecosystem.

### openxlsx package

- The **`openxlsx`** package is the dedicated tool for reading and writing data from **Excel files** (`.xlsx`) without needing external dependencies like Java. It is widely used for working with multi-sheet workbooks, applying styles, and handling complex Excel features, offering functions like `read.xlsx()` for import. 

## Why use the `rio` and `here` packages instead of base R?

### rio package

- Is highly useful because it provides a single, **universal function** for importing virtually any data file format.

- The `import()` function automatically **detects the file type** based on the file extension (e.g., `.csv`, `.xlsx`, `.RData`, `.rds`, `.xpt`, `.sas7bdat`, `.dta`, `.sav`, `.json,.txt`, `.dat`) and uses the appropriate specialized package (`readr`,`haven`, and `openxlsx`) to import it.

### here package

- The package **here** and its function `here()` make it easy to tell R where to find and to save your files - in essence, it builds file paths.

- Using **here()** works with you **R Project** to eliminate path problems -\> Relative paths = good for sharing; Absolute paths = Bad for sharing

  - **Bad:** `"/Users/mwegener/Desktop/maybe_here/data.csv"`

  - **Good:** `here("raw_data", "data.csv")`

- This is useful when the R project may be shared or accessed by multiple people/computers

### Confirm Your Working Directory

Run this to make sure you are currently in the correct R project:

```{r}
# library(here)
here()
```

::: callout-note
## If you are not in the correct project

You can set the project root manually — replace the path with your actual folder location:

``` r
# set_here("path/to/your/desired/folder")
```
:::

### Import the Data Files

Import the raw counts data

```{r}
# library(rio), library(here)
raw_data <- import(here("raw_data", "GSE113754_GeneLevel_Raw_data.csv"), header = TRUE)
```

Import the metadata

```{r}
# library(rio), library(here)
raw_metadata <- import(here("raw_data", "GSE113754_filtered_metadata.csv"), header = TRUE)
```

------------------------------------------------------------------------

## Data cleaning and validation using the Tidyverse

The **tidyverse** is a collection of R packages designed around shared principles for efficient, readable data analysis. Install all packages at once with `install.packages("tidyverse")` and load with `library(tidyverse)`.

### **Core Philosophy:**

- **Tidy data structure**: Each variable = column, each observation = row, each value = cell

- **Consistent syntax**: Predictable function behavior across all packages

- **Human-centered design**: Intuitive functions that reduce learning curve

- **Pipe workflows**: Use `%>%` (`|>`) to chain operations for readable, step-by-step analysis

### **Key Benefits:**

- **Uniform data structures** using data frames/tibbles

- **Reproducible analysis** with version control emphasis

- **Clean naming conventions** (lowercase_with_underscores)

- **Active community** support and best practices

- **Functional pipelines** that transform data through logical steps

### Core Functions

| Function | Utility | Package |
|----------------------------|--------------------------|------------------|
| `%>%` | “Pipe” (pass data from one function to the next) | magrittr |
| `mutate()` | Create, transform, and redefine columns | dplyr |
| `select()` | Keep, remove, select, or rename columns | dplyr |
| `rename()` | Rename columns | dplyr |
| `clean_names()` | Standardize column-name syntax | janitor |
| `as.character()`, `as.numeric()`, `as.Date()`, *etc.* | Convert the class of a column | base R |
| `across()` | Transform multiple columns at once | dplyr |
| *tidyselect* helpers | Use logic to select columns | tidyselect |
| `filter()` | Keep certain rows | dplyr |
| `distinct()` | De-duplicate rows | dplyr |
| `rowwise()` | Perform operations within each row | dplyr |
| `add_row()` | Add rows manually | tibble |
| `arrange()` | Sort rows | dplyr |
| `recode()` | Recode values in a column | dplyr |
| `case_when()` | Recode values with complex logical criteria | dplyr |
| `replace_na()`, `na_if()`, `coalesce()` | Handle missing values | tidyr |
| `age_categories()`, `cut()` | Create categorical groups from a numeric column | epikit / base R |
| `match_df()` | Recode/clean values using a data dictionary | matchmaker |
| `which()` | Apply logical criteria; return indices | base R |

## Let's start with the metadata file 'raw_metadata'

### First, Inspect the Data

#### Inspecting data helps identify potential problems before analysis and plan your data cleaning strategy by verifying:

- Data imported correctly

- Variable types are appropriate

- Value formatting is consistent

- Missing value patterns

- Duplicate records

- Data set structure and dimensions

- **If column names are clean (concise, no spaces, no blanks, no dots, no caps)**

Physically look at the data you just imported:

```{r}
view(raw_metadata)
```

Use `head()` to see the first 6 rows, column names, and data types:

```{r}
head(raw_metadata)
```

Use `str()` to see the structure, data types, dimensions, and how data are organized:

```{r}
str(raw_metadata)
```

::: callout-tip
## EXTRA: Count unique values in a column

We can find the unique values within a variable and their counts using `count()` — similar to the `unique()` and `table()` exercises from the Data Structures section:

```{r}
raw_metadata %>% count(genotype)
```
:::

### What Changes Do We Need to Make?

1.  Fix column names
2.  Rename `characteristic` to `condition`
3.  Clean up values under `condition`
4.  Clean up values under `genotype`
5.  Combine `condition` and `genotype` into a new column `geno_cond`
6.  Keep columns in this order: `sample_id`, `genotype`, `condition`, `geno_cond`
7.  Make values under `sample_id` lowercase

::: callout-tip
## Working with the pipe `%>%`

Each step below builds on the previous one using the pipe operator `%>%`. You can run the code up to any `%>%` to inspect the result at that point — this is a great way to debug and understand how the data changes at each step.
:::

### Next, clean the meta data file

#### Step 1: Fix All Column Names

`clean_names()` standardizes all column names to lowercase with underscores:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names()
```

#### Step 2: Rename Columns

Rename column `V1` to `sample_id` and `characteristics` to `condition`:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names() %>%
  rename(sample_id = V1, condition = characteristics)
```

#### Step 3: Recode Values in `condition`

- If the value includes `"homecage"` → change to `"normal"`
- If the value includes `"sleep deprivation"` → change to `"sleep_deprived"`

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names() %>%
  rename(sample_id = v1, condition = characteristics) %>%
  mutate(
    condition = case_when(
      grepl("homecage", condition)        ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    ))
```

::: callout-note
## Base R inside the Tidyverse

Notice that `grepl()` is a base R function — you can use base R freely within Tidyverse pipelines. This is the same function used in the Data Structures section with `ifelse()`.
:::

#### Step 4: Recode Values in `genotype`

- If the value includes `"WT"` → change to `"WT"`
- If the value includes `"shank3"` → change to `"shank3"`

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names() %>%
  rename(sample_id = v1, condition = characteristics) %>%
  mutate(
    condition = case_when(
      grepl("homecage", condition)          ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    )) %>%
  mutate(
    genotype = case_when(
      grepl("WT",     genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype
    ))
```

#### Step 5: Create `geno_cond` Column

Combine `genotype` and `condition` into a new variable:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names() %>%
  rename(sample_id = v1, condition = characteristics) %>%
  mutate(
    condition = case_when(
      grepl("homecage", condition)          ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    )) %>%
  mutate(
    genotype = case_when(
      grepl("WT",     genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype
    )) %>%
  mutate(geno_cond = paste0(genotype, condition))
```

#### Step 6: Select and Reorder Columns

Keep only the columns we need in the correct order:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names() %>%
  rename(sample_id = v1, condition = characteristics) %>%
  mutate(
    condition = case_when(
      grepl("homecage", condition)          ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    )) %>%
  mutate(
    genotype = case_when(
      grepl("WT",     genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype
    )) %>%
  mutate(geno_cond = paste0(genotype, condition)) %>%
  select(sample_id, genotype, condition, geno_cond)
```

#### Step 7: Lowercase `sample_id`

Make `sample_id` values lowercase — this will be important for harmonizing with the raw counts data in the next section:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_meta_data <- raw_metadata %>%
  clean_names() %>%
  rename(sample_id = v1, condition = characteristics) %>%
  mutate(
    condition = case_when(
      grepl("homecage", condition)          ~ "normal",
      grepl("sleep deprivation", condition) ~ "sleep_deprived",
      TRUE ~ condition
    )) %>%
  mutate(
    genotype = case_when(
      grepl("WT",     genotype, ignore.case = TRUE) ~ "WT",
      grepl("shank3", genotype, ignore.case = TRUE) ~ "shank3",
      TRUE ~ genotype
    )) %>%
  mutate(geno_cond = paste0(genotype, condition)) %>%
  select(sample_id, genotype, condition, geno_cond) %>%
  mutate(sample_id = str_to_lower(sample_id))
```

------------------------------------------------------------------------

## Now, let's move on to the 'raw_data' file

### First, Inspect the Data

```{r}
view(raw_data)
```

```{r}
head(raw_data)
```

```{r}
str(raw_data)
```

### What Issues Did You Find?

- Column 1 has no label and contains row numbers inserted in error
- Column 2 has no label
- Column 3 (`gene symbol`) has a space in its name
- All columns beginning with `GSM` are uppercase

::: callout-tip
## EXTRA: Search for NA values

Use `filter()`, `if_any()`, and `everything()` with `is.na()` to search all rows and columns for missing values:

```{r}
#library(tidyverse), library(dplyr)
raw_data %>%
  filter(if_any(everything(), is.na))
```

The NA value identified here will be addressed in the cleaning steps below.
:::

### Next, we will clean the 'raw_data' file

#### `clean_names()`

Gives columns 1 and 2 auto-generated labels, replaces spaces in column names with underscores, and converts all uppercase letters to lowercase:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_data <- raw_data %>%
  clean_names()
```

#### `rename()`

Change the auto-generated name for column 2 (`v1`) to `gene_id`:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_data <- raw_data %>%
  clean_names() %>%
  rename(gene_id = v1)
```

#### `mutate()` — Handle the NA Value

Replace the NA value under `gsm3119015` with `0`:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_data <- raw_data %>%
  clean_names() %>%
  rename(gene_id = v1) %>%
  mutate(gsm3119015 = replace_na(gsm3119015, 0))
  # drop_na() %>%  # alternative: drop all rows with at least one NA
```

::: callout-important
## A note on handling missing values

Replacing NA with `0` is used here for demonstration purposes. In practice, the right approach depends on your data and analysis — options include removing rows, leaving the NA, or calculating an imputed value. Always think carefully before replacing missing data.
:::

#### `mutate()` — Create a Combined Gene ID Column

Create a new variable that combines `gene_id` and `gene_symbol` separated by `|`. Notice that `paste()` from base R works seamlessly inside a Tidyverse pipeline:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_data <- raw_data %>%
  clean_names() %>%
  rename(gene_id = v1) %>%
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|"))
```

#### `select()`

Make `gene_id_symbol` the first column, keep everything else in the same order using `everything()`, and remove `v1`, `gene_symbol`, and `gene_id` using the `-` operator:

```{r}
# library(janitor), library(tidyverse), library(dplyr)
clean_data <- raw_data %>%
  clean_names() %>%
  rename(gene_id = v1) %>%
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|")) %>%
  select(gene_id_symbol, everything(), -gene_symbol, -gene_id)
```

------------------------------------------------------------------------

## EXTRA: Harmonizing the Two Data Frames

These extra steps align the structure of `clean_data` with `clean_meta_data` so they are ready for downstream analysis such as differential gene expression.

### Extra 1: Set Row IDs

Use `column_to_rownames()` from the `tibble` package to convert the first column from a data variable into row identifiers:

```{r}
# library(janitor), library(tidyverse), library(dplyr), library(tibble)
clean_data <- raw_data %>%
  clean_names() %>%
  rename(gene_id = v1) %>%
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|")) %>%
  select(gene_id_symbol, everything(), -gene_symbol, -gene_id) %>%
  column_to_rownames(var = colnames(.)[1])
```

### Extra 2: Reorder Columns to Match Metadata

Reorder the columns in `clean_data` to match the order of `sample_id` values in `clean_meta_data`:

```{r}
# library(janitor), library(tidyverse), library(dplyr), library(tibble)
clean_data <- raw_data %>%
  clean_names() %>%
  rename(gene_id = v1) %>%
  mutate(gsm3119015 = replace_na(gsm3119015, 0)) %>%
  mutate(gene_id_symbol = paste(gene_id, gene_symbol, sep = "|")) %>%
  select(gene_id_symbol, everything(), -gene_symbol, -gene_id) %>%
  column_to_rownames(var = colnames(.)[1]) %>%
  select(all_of(clean_meta_data$sample_id))
```

### Extra 3: Verify Harmonization

Run a quick check to confirm that the column names in `clean_data` and the `sample_id` values in `clean_meta_data` match exactly:

```{r}
setequal(clean_meta_data$sample_id, colnames(clean_data))
```

------------------------------------------------------------------------

## Save Cleaned Data

Save `clean_data` and `clean_meta_data` locally using `here()` and 'export()\`:

```{r}
# Saves with row labels (use read.csv to reload with row labels intact)
export(clean_data, here("processed_data", "clean_data_label.csv"))
```

```{r}
# Saves without row labels (standard clean CSV)
export(clean_data, here("processed_data", "clean_data.csv"))
```

```{r}
# Save clean metadata
export(clean_meta_data, here("processed_data", "clean_meta_data.csv"))
```

------------------------------------------------------------------------

::: callout-note
## Can You...?

- [x] Use rio::import() and rio::export() to read and write diverse file formats without manually switching packages?
- [x] Explain what `here()` does and why it improves reproducibility?
- [x] Describe at least two differences between `read_csv()` and `read.csv()`?
- [x] Explain what a tibble is and how it differs from a data frame?
- [x] Use `%>%` to chain multiple data cleaning steps together?
- [x] Use `clean_names()` to standardize column names automatically?
- [x] Use `rename()`, `mutate()`, and `case_when()` to recode values?
- [x] Use `select()` and `everything()` to reorder and drop columns?
- [x] Replace NA values using `replace_na()`?
- [x] Explain what data harmonization means and verify it with `setequal()`?
- [x] Choose between `write_csv()` and `write.csv()` based on whether you need row labels?
:::