---
title: "Finding Differentially Expressed Genes with DESeq2"
subtitle: "Workshop: Shank3 vs WT Genotype Comparison"
author: "Sofia Fertuzinhos, PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    theme: flatly
    highlight-style: github
    self-contained: true
execute:
  warning: false
  message: false
---

# 🧬 Introduction

In this workshop we will use **DESeq2** to identify differentially expressed genes between two genotypes: **Shank3** and **Wild Type (WT)** mice.

We will work through a complete RNA-seq analysis pipeline:

1.  Load and inspect the data\
2.  Set up the DESeq2 object\
3.  Run the analysis\
4.  Perform quality control\
5.  Extract and interpret results\
6.  Export tables and plots

## Project Structure

This script assumes the following R project structure:

```{plaintext}
ShankProject/
├── renv/
├── renv.lock                 ← locked package versions (tracked by Git)
├── .Rprofile                 ← auto-activates 
├── processed_data/
│   └── Shank3_metadata_clean.csv
│   └── Shank3_rawCounts_clean.csv
├── raw_data/
    ├── GSE113754_GeneLevel_Raw_data.csv
│   └── GSE113754_filtered_metadata.csv
├── scripts/
│   ├── 07_DESeq2_pipeline.qmd          
│   └── 07_setup_script.r     ← ✅ installation script (run ONCE)
├── outputs/
└── README.md
```

## Environment & Reproducibility

This project uses **renv** for package management and is version-controlled with **Git/GitHub**.

```{r}
### To restore the exact package environment on a new machine:
renv::restore()

### To snapshot new packages you install:
renv::snapshot()
```

# Part 1 — Setup

## 1.1 Load Required Packages

```{r}
library(here)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(tibble)
library(pheatmap)
library(RColorBrewer)
library(apeglm)     # used by lfcShrink() in Part 7
```

::: callout-note
`apeglm` is loaded here even though we never call it directly — `lfcShrink()`
reaches it via `type = "apeglm"`. Because that's a string, `renv` cannot see
the dependency when it scans your code, so without this `library()` call
`renv::snapshot()` leaves `apeglm` out of the lockfile and Part 7 fails for
anyone who restores from it. Same reasoning for `tibble`, used as
`tibble::rownames_to_column()`.
:::

## 1.2 Define File Paths

Why use here::here()? Using here() makes your paths relative to the project root, so the script works on any machine regardless of where the project folder is saved. This is especially important when sharing code via GitHub.
```{r}
# | label: define-paths

# Input paths
path_counts <- here("processed_data", "Shank3_rawCounts_clean.csv")
path_metadata <- here("processed_data", "Shank3_metadata_clean.csv")

# Output paths
path_figures <- here("outputs", "figures")
path_tables <- here("outputs", "tables")

# Create output folders if they do not exist yet
dirs_to_create <- c(path_figures, path_tables)

for (dir in dirs_to_create) {
if (!dir.exists(dir)) {
dir.create(dir, recursive = TRUE)
cat("✅ Created:", dir, "\\n")
}
}
```

# 📂 Part 2 — Load & Inspect the Data

## 2.1 Load Raw Counts

```{r}
clean_counts <- read.csv(path_counts, header = TRUE, row.names = 1)
```

Quick inspection

```{r}
cat("Dimensions: ", nrow(clean_counts), "genes x", ncol(clean_counts), "samples\n") 
head(counts[ , 1:5]) # show first 5 samples only 2.2 Load Sample Metadata 
```

## 2.2 Load Metadata

```{r}
clean_info <- read.csv(path_metadata, header = TRUE, stringsAsFactors = TRUE)
```

Quick inspection

```{r}
print(clean_info)
```

Check: Do columns in rawc match rows in info? The column names of the counts matrix must match the row names (or sample identifiers) in the metadata, in the same order. This is a very common source of errors. :::

Important: Verify sample names match between counts and metadata

```{r}
all(colnames(clean_counts) == clean_info$sample_id)

# Should return TRUE — if FALSE, reorder before proceeding 2.3 Subset to Genotype Comparison Only
```

# ✂️ Part 3 - Subset dataset to a simple two-group comparison

Why subset? The full dataset contains samples from two conditions (normal and sleep-deprived) and two genotypes (WT and Shank3).

For this workshop we focus on a simple two-group comparison: Shank3 vs WT, using only the normal condition samples. This keeps the design formula simple and the results easier to interpret. :::

### 3.1 - Keep only samples from the normal (baseline) condition

```{r}
info_sub <- clean_info %>% filter(condition == "normal")
```

### 3.2 - Subset counts to matching samples

```{r}
counts_sub <- clean_counts[ , info_sub$sample_id]
```

### 3.3 - Drop unused factor levels after subsetting

```{r}
info_sub$genotype <- droplevels(info_sub$genotype)
```

### 3.4 - Set WT as the reference level (baseline for comparison)

```{r}
info_sub$genotype <- relevel(info_sub$genotype, ref = "WT")
```

### 3.5 Confirm subset dimensions

```{r}
cat("Subsetted dataset:", nrow(counts_sub), "genes x", ncol(counts_sub), "samples\n")
print(info_sub)
```

# 🏗️ Part 4 — Build the DESeq2 Object

### 4.1 Create DESeqDataSet

```{r}
dds <- DESeqDataSetFromMatrix( countData = counts_sub, # genes x samples count matrix 
                               colData = info_sub, # sample metadata 
                               design = ~ genotype # simple two-group design
                               )
```

#### Understanding the design formula:

**genotype** tells DESeq2 to model gene expression as a function of genotype.

With WT set as the reference, all results will be expressed as: Shank3 relative to WT


### 4.2 Pre-filter Low-Count Genes plaintext

Keep genes with at least 10 counts in at least 3 samples (adjust thresholds based on your smallest group size)

```{r}
keep <- rowSums(counts(dds) >= 10) >= 3 
dds <- dds[keep, ]

cat("Genes remaining after filtering:", nrow(dds), "\n") 
```

# 🔬 Part 5 — Run DESeq2

```{r}
dds <- DESeq(dds)
```

What does DESeq() do? Three steps happen automatically:

Size factor estimation — normalizes for differences in sequencing depth Dispersion estimation — models count variability across samples Model fitting — fits a negative binomial GLM and tests each gene

# 📊 Part 6 — Quality Control

### 6.1 Transform Counts for Visualization

vst = Variance Stabilizing Transformation

Faster than rlog() for larger datasets; both are suitable for QC

```{r}
vsd <- vst(dds, blind = TRUE)
```

blind = TRUE vs blind = FALSE

blind = TRUE — transformation ignores the design (used for QC/exploration) blind = FALSE — uses the design (used for downstream visualization of results)

### 6.2 PCA Plot

What to look for in the PCA Samples should cluster by genotype along PC1 (the axis of greatest variation) Any outlier samples should be investigated before proceeding If genotype is not the dominant source of variation, review your metadata

```{r}
pca_data <- plotPCA(vsd, intgroup = "genotype", returnData = TRUE) 

pct_var <- round(100 * attr(pca_data, "percentVar"))

pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = genotype, label = name)) + 
            geom_point(size = 4, alpha = 0.8) + 
            geom_text(vjust = -0.8, size = 3) + 
            scale_color_manual(values = c("WT" = "#2196F3", "Shank3" = "#F44336")) + 
            labs( title = "PCA: Shank3 vs WT (normal condition)", x = paste0("PC1: ", pct_var[1], "% variance"), y = paste0("PC2: ", pct_var[2], "% variance") ) + 
            theme_classic(base_size = 13) + 
            theme(legend.position = "right")

pca_plot

# Save
ggsave(filename = file.path(path_figures, "PCA_Shank3vsWT.png"), plot = pca_plot, width = 7, height = 5, dpi = 300)
```

### 6.3 Sample Distance Heatmap

```{r}
samp_dists <- dist(t(assay(vsd))) 
samp_mat <- as.matrix(samp_dists)

rownames(samp_mat) <- colnames(samp_mat) <- paste(colData(vsd)$genotype, colData(vsd)$sample_id, sep = " | ")

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(samp_mat, col = colors, 
         clustering_distance_rows = samp_dists, 
         clustering_distance_cols = samp_dists, 
         main = "Sample-to-Sample Distances", fontsize = 10, 
         filename = file.path(path_figures, "SampleDistances_Shank3vsWT.png"))

dev.new()
pheatmap(samp_mat, col = colors,
         clustering_distance_rows = samp_dists,
         clustering_distance_cols = samp_dists,
         main = "Sample-to-Sample Distances",
         fontsize = 10)
```

# 📈 Part 7 — Extract & Interpret Results

### 7.1 Check Available Coefficients plaintext #\| label: results-names

```{r}
resultsNames(dds) 
# Expected output: 
# [1] "Intercept" 
# [2] "genotype_Shank3_vs_WT" 
```

### 7.2 Extract Results: Shank3 vs WT plaintext

```{r}
#| label: extract-results

res <- results(dds, name = "genotype_Shank3_vs_WT", alpha = 0.05) # sets the FDR threshold for independent filtering

summary(res)
```

::: callout-note
Understanding the results columns Column Meaning baseMean Average normalized count across all samples log2FoldChange Log2 ratio of Shank3 / WT lfcSE Standard error of the log2 fold change stat Wald test statistic pvalue Raw p-value padj Adjusted p-value (FDR) — use this for significance testing
:::

### 7.3 Shrink Fold Changes (Recommended)

```{r}
#| label: lfc-shrink

# lfcShrink improves fold change estimates for genes with low counts
# This is recommended for ranking and visualization — NOT for hypothesis testing

res_shrunk <- lfcShrink(dds, coef = "genotype_Shank3_vs_WT", type = "apeglm")

summary(res_shrunk)
```

::: callout-tip
Why shrink fold changes? Genes with low counts can have artificially extreme fold changes due to high variability. lfcShrink() pulls these estimates towards zero, making volcano plots and ranked gene lists more reliable.

Use res_shrunk for visualization. Use res for hypothesis testing.
:::

### 7.4 MA Plot

```{r}
# | label: ma-plot 
# | fig-width: 7
# | fig-height: 5 
# | fig-cap: "MA plot showing log2 fold change (Shank3/WT) versus mean expression. Significant genes (padj < 0.05) are highlighted in blue."

# Save to file and display

png(file.path(path_figures, "MAplot_Shank3vsWT.png"), width = 700, height = 500, res = 120) 

plotMA(res_shrunk, ylim = c(-5, 5), 
                    main = "MA Plot: Shank3 vs WT", 
                    alpha = 0.1, 
                    colSig = "#F44336", 
                    colNonSig = "grey60") 

dev.off()

# Display inline

plotMA(res_shrunk, ylim = c(-5, 5), 
                    main = "MA Plot: Shank3 vs WT", 
                    alpha = 0.1, 
                    colSig = "#F44336", 
                    colNonSig = "grey60") 
```

### 7.5 Volcano Plot

```{r}
res_df <- as.data.frame(res_shrunk) %>%
  tibble::rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>%
  mutate(
    significance = case_when(
      padj < 0.1 & log2FoldChange > 0  ~ "Up in Shank3",
      padj < 0.1 & log2FoldChange < 0 ~ "Down in Shank3",
      TRUE                            ~ "Not significant"
    )
  )

deg_counts <- table(res_df$significance)
deg_counts

volcano_plot <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c(
    "Up in Shank3" = "#F44336",
    "Down in Shank3" = "#2196F3",
    "Not significant" = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  labs(
    title = "Volcano Plot: Shank3 vs WT",
    subtitle = paste0(
      "Up: ", deg_counts["Up in Shank3"],
      " | Down: ", deg_counts["Down in Shank3"]
    ),
    x = "Log2 Fold Change (Shank3 / WT)",
    y = "-Log10 Adjusted P-value",
    color = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

volcano_plot
ggsave(file.path(path_figures, "Volcano_Shank3vsWT.png"), plot = volcano_plot, width = 8, height = 6, dpi = 300)
```

# 💾 Part 8 — Export Results

### 8.1 Filter Significant DEGs

```{r}
# | label: filter-degs

# All results (sorted by adjusted p-value)

res_all <- as.data.frame(res_shrunk) %>% tibble::rownames_to_column("gene") %>% arrange(padj)

# Significant DEGs only (padj < 0.05, |LFC| > 1)

res_sig <- res_all %>% filter(padj < 0.05, abs(log2FoldChange) > 1)

cat("Total DEGs (padj < 0.05, |LFC| > 1):", nrow(res_sig), "\n")
head(res_sig)
```

### 8.2 Save Results Tables

```{r}
# | label: save-tables

# Full results table

write.csv(res_all, file = file.path(path_tables, "DESeq2_Shank3vsWT_allGenes.csv"), row.names = FALSE)

# Significant DEGs only

write.csv(res_sig, file = file.path(path_tables, "DESeq2_Shank3vsWT_sigDEGs.csv"), row.names = FALSE)

# Normalised counts

norm_counts <- counts(dds, normalized = TRUE) 
write.csv(norm_counts, file = file.path(path_tables, "normCounts_Shank3vsWT.csv"), row.names = TRUE)

cat("✅ Tables saved to outputs/tables \n") 
```

# 🔁 Part 9 — Reproducibility Session Information
```{r}
# | label: session-info 
# | code-fold: true
sessionInfo()
```

::: callout-tip
Saving your renv snapshot After the workshop, lock your package versions so results are reproducible:

```{r}
 renv::snapshot() 
```

Commit the updated renv.lock to GitHub:

```{bash}
git add renv.lock 
git commit -m "update renv snapshot after workshop" 
git push 
```
:::

# 📚 Summary

| Step | What we did |
|------------------------------------|------------------------------------|
| **Load data** | Read counts matrix and sample metadata |
| **Subset** | Kept only normal-condition samples for a clean 2-group comparison |
| **Pre-filter** | Removed genes with very low counts |
| **DESeq2** | Ran normalisation, dispersion estimation and Wald test |
| **QC** | PCA and sample distance heatmap |
| **Results** | Extracted Shank3 vs WT comparison with LFC shrinkage |
| **Visualize** | MA plot, volcano plot, top gene counts |
| **Export** | Saved tables to `/tables`, plots to `/images`, counts to `/processed_data` |

::: callout-note
## 🔭 Next Steps (beyond this workshop)

- Gene Ontology (GO) enrichment analysis with `clusterProfiler` and `fgsea`
- Heatmap of top DEGs with `pheatmap`
- More complex designs: adding condition and interaction terms
:::

------------------------------------------------------------------------

