---
title: "Finding Differentially Expressed Genes with DESeq2"
subtitle: "Workshop: Shank3 vs WT Genotype Comparison"
author: "Sofia Fertuzinhos, PhD"
sidebar: arcs04
format:
  html:
    toc-depth: 3
    code-tools: true
execute:
  warning: false
  message: false
  freeze: auto
---

## 🧬 Introduction

In this workshop we will use **DESeq2** to identify differentially expressed genes between two genotypes: **Shank3** and **Wild Type (WT)** mice.

We will work through a complete RNA-seq analysis pipeline:

1.  Load and inspect the data
2.  Set up the DESeq2 object
3.  Run the analysis
4.  Perform quality control
5.  Extract and interpret results
6.  Export tables and plots

### Project Structure

This script assumes the following R project structure:

```
ShankProject/
├── renv/
├── renv.lock                 ← locked package versions (tracked by Git)
├── .Rprofile                 ← auto-activates renv
├── processed_data/
│   ├── Shank3_metadata_clean.csv
│   └── Shank3_rawCounts_clean.csv
├── raw_data/
│   ├── GSE113754_GeneLevel_Raw_data.csv
│   └── GSE113754_filtered_metadata.csv
├── scripts/
│   ├── 07-deseq2_pipeline.qmd
│   └── 07-setup_script.r     ← ✅ installation script (run ONCE)
├── outputs/
└── README.md
```

### Environment & Reproducibility

This project uses **renv** for package management and is version-controlled with **Git/GitHub**.

```{r}
#| eval: false

# To restore the exact package environment on a new machine:
renv::restore()

# To snapshot new packages you install:
renv::snapshot()
```

## Part 1 — Setup

### 1.1 Load Required Packages

```{r}
#| label: load-packages

library(here)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(tibble)
library(pheatmap)
library(RColorBrewer)
library(apeglm)     # used by lfcShrink() in Part 7
```

::: callout-note
`apeglm` is loaded here even though we never call it directly — `lfcShrink()`
reaches it via `type = "apeglm"`. Because that's a string, `renv` cannot see
the dependency when it scans your code, so without this `library()` call
`renv::snapshot()` leaves `apeglm` out of the lockfile and Part 7 fails for
anyone who restores from it. Same reasoning for `tibble`, used as
`tibble::rownames_to_column()`.
:::

### 1.2 Define File Paths

Why use `here::here()`? Using `here()` makes your paths relative to the project root, so the script works on any machine regardless of where the project folder is saved. This is especially important when sharing code via GitHub.

```{r}
#| label: define-paths

# In your own ShankProject the cleaned data sits at <project root>/processed_data,
# which is the first branch below. This page is published from the ARCS website
# repository, where the same two files ship alongside the workshop materials —
# the fallback keeps the published version reproducible without changing the
# path you would write in your own project.
dir_data <- if (dir.exists(here("processed_data"))) {
  here("processed_data")
} else {
  here("workshops", "arcs_04", "_materials", "processed_data")
}

dir_out <- if (dir.exists(here("processed_data"))) {
  here("outputs")
} else {
  here("workshops", "arcs_04", "_materials", "outputs")
}

# Input paths
path_counts   <- file.path(dir_data, "Shank3_rawCounts_clean.csv")
path_metadata <- file.path(dir_data, "Shank3_metadata_clean.csv")

# Output paths
path_figures <- file.path(dir_out, "figures")
path_tables  <- file.path(dir_out, "tables")

# Create output folders if they do not exist yet
for (dir in c(path_figures, path_tables)) {
  if (!dir.exists(dir)) {
    dir.create(dir, recursive = TRUE)
    cat("✅ Created:", dir, "\n")
  }
}
```

## 📂 Part 2 — Load & Inspect the Data

### 2.1 Load Raw Counts

```{r}
#| label: load-counts

clean_counts <- read.csv(path_counts, header = TRUE, row.names = 1)
```

Quick inspection:

```{r}
#| label: inspect-counts

cat("Dimensions: ", nrow(clean_counts), "genes x", ncol(clean_counts), "samples\n")
head(clean_counts[, 1:5])   # show first 5 samples only
```

### 2.2 Load Metadata

```{r}
#| label: load-metadata

clean_info <- read.csv(path_metadata, header = TRUE, stringsAsFactors = TRUE)
```

Quick inspection:

```{r}
#| label: inspect-metadata

print(clean_info)
```

::: callout-important
## Do the columns in the counts matrix match the rows in the metadata?

The column names of the counts matrix must match the sample identifiers in the
metadata, **in the same order**. This is a very common source of errors — check
it before you build anything downstream.
:::

```{r}
#| label: check-sample-order

all(colnames(clean_counts) == clean_info$sample_id)

# Should return TRUE — if FALSE, reorder before proceeding
```

## ✂️ Part 3 — Subset to a Simple Two-Group Comparison

Why subset? The full dataset contains samples from two conditions (normal and sleep-deprived) and two genotypes (WT and Shank3).

For this workshop we focus on a simple two-group comparison: **Shank3 vs WT**, using only the *normal* condition samples. This keeps the design formula simple and the results easier to interpret.

### 3.1 Keep only samples from the normal (baseline) condition

```{r}
#| label: subset-metadata

info_sub <- clean_info %>% filter(condition == "normal")
```

### 3.2 Subset counts to matching samples

```{r}
#| label: subset-counts

counts_sub <- clean_counts[, info_sub$sample_id]
```

### 3.3 Drop unused factor levels after subsetting

```{r}
#| label: droplevels

info_sub$genotype <- droplevels(info_sub$genotype)
```

### 3.4 Set WT as the reference level (baseline for comparison)

```{r}
#| label: relevel

info_sub$genotype <- relevel(info_sub$genotype, ref = "WT")
```

### 3.5 Confirm subset dimensions

```{r}
#| label: confirm-subset

cat("Subsetted dataset:", nrow(counts_sub), "genes x", ncol(counts_sub), "samples\n")
print(info_sub)
```

## 🏗️ Part 4 — Build the DESeq2 Object

### 4.1 Create DESeqDataSet

```{r}
#| label: build-dds

dds <- DESeqDataSetFromMatrix(
  countData = counts_sub,   # genes x samples count matrix
  colData   = info_sub,     # sample metadata
  design    = ~ genotype    # simple two-group design
)
```

::: callout-note
## Understanding the design formula

`~ genotype` tells DESeq2 to model gene expression as a function of genotype.

With WT set as the reference, all results will be expressed as **Shank3 relative to WT**.
:::

### 4.2 Pre-filter Low-Count Genes

Keep genes with at least 10 counts in at least 3 samples (adjust thresholds based on your smallest group size).

```{r}
#| label: prefilter

keep <- rowSums(counts(dds) >= 10) >= 3
dds  <- dds[keep, ]

cat("Genes remaining after filtering:", nrow(dds), "\n")
```

## 🔬 Part 5 — Run DESeq2

```{r}
#| label: run-deseq

dds <- DESeq(dds)
```

::: callout-note
## What does `DESeq()` do?

Three steps happen automatically:

1.  **Size factor estimation** — normalizes for differences in sequencing depth
2.  **Dispersion estimation** — models count variability across samples
3.  **Model fitting** — fits a negative binomial GLM and tests each gene
:::

## 📊 Part 6 — Quality Control

### 6.1 Transform Counts for Visualization

`vst` = Variance Stabilizing Transformation. It is faster than `rlog()` for larger datasets; both are suitable for QC.

```{r}
#| label: vst

vsd <- vst(dds, blind = TRUE)
```

::: callout-note
## `blind = TRUE` vs `blind = FALSE`

- **`blind = TRUE`** — the transformation ignores the design (used for QC and exploration)
- **`blind = FALSE`** — the transformation uses the design (used for downstream visualization of results)
:::

### 6.2 PCA Plot

::: callout-tip
## What to look for in the PCA

- Samples should cluster by genotype along PC1 (the axis of greatest variation)
- Any outlier samples should be investigated before proceeding
- If genotype is not the dominant source of variation, review your metadata
:::

```{r}
#| label: pca-plot
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "PCA of variance-stabilized counts, coloured by genotype."

pca_data <- plotPCA(vsd, intgroup = "genotype", returnData = TRUE)

pct_var <- round(100 * attr(pca_data, "percentVar"))

pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = genotype, label = name)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text(vjust = -0.8, size = 3) +
  scale_color_manual(values = c("WT" = "#2196F3", "Shank3" = "#F44336")) +
  labs(
    title = "PCA: Shank3 vs WT (normal condition)",
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance")
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "right")

pca_plot

# Save
ggsave(
  filename = file.path(path_figures, "PCA_Shank3vsWT.png"),
  plot = pca_plot, width = 7, height = 5, dpi = 300
)
```

### 6.3 Sample Distance Heatmap

```{r}
#| label: distance-heatmap
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Euclidean distances between samples, computed on the variance-stabilized counts."

samp_dists <- dist(t(assay(vsd)))
samp_mat   <- as.matrix(samp_dists)

rownames(samp_mat) <- colnames(samp_mat) <-
  paste(colData(vsd)$genotype, colData(vsd)$sample_id, sep = " | ")

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

# Display inline
pheatmap(samp_mat, col = colors,
         clustering_distance_rows = samp_dists,
         clustering_distance_cols = samp_dists,
         main = "Sample-to-Sample Distances",
         fontsize = 10)

# Save to file
pheatmap(samp_mat, col = colors,
         clustering_distance_rows = samp_dists,
         clustering_distance_cols = samp_dists,
         main = "Sample-to-Sample Distances", fontsize = 10,
         filename = file.path(path_figures, "SampleDistances_Shank3vsWT.png"))
```

## 📈 Part 7 — Extract & Interpret Results

### 7.1 Check Available Coefficients

```{r}
#| label: results-names

resultsNames(dds)
```

### 7.2 Extract Results: Shank3 vs WT

```{r}
#| label: extract-results

# alpha sets the FDR threshold used for independent filtering
res <- results(dds, name = "genotype_Shank3_vs_WT", alpha = 0.05)

summary(res)
```

::: callout-note
## Understanding the results columns

| Column | Meaning |
|-------------------|-----------------------------------------------------|
| `baseMean` | Average normalized count across all samples |
| `log2FoldChange` | Log2 ratio of Shank3 / WT |
| `lfcSE` | Standard error of the log2 fold change |
| `stat` | Wald test statistic |
| `pvalue` | Raw p-value |
| `padj` | Adjusted p-value (FDR) — **use this for significance testing** |
:::

### 7.3 Shrink Fold Changes (Recommended)

```{r}
#| label: lfc-shrink

# lfcShrink improves fold change estimates for genes with low counts.
# This is recommended for ranking and visualization — NOT for hypothesis testing.

res_shrunk <- lfcShrink(dds, coef = "genotype_Shank3_vs_WT", type = "apeglm")

summary(res_shrunk)
```

::: callout-tip
## Why shrink fold changes?

Genes with low counts can have artificially extreme fold changes due to high
variability. `lfcShrink()` pulls these estimates towards zero, making volcano
plots and ranked gene lists more reliable.

Use `res_shrunk` for **visualization**. Use `res` for **hypothesis testing**.
:::

### 7.4 MA Plot

```{r}
#| label: ma-plot
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "MA plot showing log2 fold change (Shank3/WT) versus mean expression. Significant genes are highlighted in red."

# Save to file
png(file.path(path_figures, "MAplot_Shank3vsWT.png"),
    width = 700, height = 500, res = 120)

plotMA(res_shrunk, ylim = c(-5, 5),
       main      = "MA Plot: Shank3 vs WT",
       alpha     = 0.1,
       colSig    = "#F44336",
       colNonSig = "grey60")

invisible(dev.off())

# Display inline
plotMA(res_shrunk, ylim = c(-5, 5),
       main      = "MA Plot: Shank3 vs WT",
       alpha     = 0.1,
       colSig    = "#F44336",
       colNonSig = "grey60")
```

### 7.5 Volcano Plot

```{r}
#| label: volcano-plot
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Volcano plot of adjusted significance against effect size."

res_df <- as.data.frame(res_shrunk) %>%
  tibble::rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>%
  mutate(
    significance = case_when(
      padj < 0.1 & log2FoldChange > 0 ~ "Up in Shank3",
      padj < 0.1 & log2FoldChange < 0 ~ "Down in Shank3",
      TRUE                            ~ "Not significant"
    )
  )

deg_counts <- table(res_df$significance)
deg_counts

volcano_plot <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c(
    "Up in Shank3"    = "#F44336",
    "Down in Shank3"  = "#2196F3",
    "Not significant" = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  labs(
    title = "Volcano Plot: Shank3 vs WT",
    subtitle = paste0(
      "Up: ", deg_counts["Up in Shank3"],
      " | Down: ", deg_counts["Down in Shank3"]
    ),
    x = "Log2 Fold Change (Shank3 / WT)",
    y = "-Log10 Adjusted P-value",
    color = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

volcano_plot

ggsave(file.path(path_figures, "Volcano_Shank3vsWT.png"),
       plot = volcano_plot, width = 8, height = 6, dpi = 300)
```

## 💾 Part 8 — Export Results

### 8.1 Filter Significant DEGs

```{r}
#| label: filter-degs

# All results (sorted by adjusted p-value)
res_all <- as.data.frame(res_shrunk) %>%
  tibble::rownames_to_column("gene") %>%
  arrange(padj)

# Significant DEGs only (padj < 0.05, |LFC| > 1)
res_sig <- res_all %>% filter(padj < 0.05, abs(log2FoldChange) > 1)

cat("Total DEGs (padj < 0.05, |LFC| > 1):", nrow(res_sig), "\n")
head(res_sig)
```

### 8.2 Save Results Tables

```{r}
#| label: save-tables

# Full results table
write.csv(res_all, file = file.path(path_tables, "DESeq2_Shank3vsWT_allGenes.csv"),
          row.names = FALSE)

# Significant DEGs only
write.csv(res_sig, file = file.path(path_tables, "DESeq2_Shank3vsWT_sigDEGs.csv"),
          row.names = FALSE)

# Normalised counts
norm_counts <- counts(dds, normalized = TRUE)
write.csv(norm_counts, file = file.path(path_tables, "normCounts_Shank3vsWT.csv"),
          row.names = TRUE)

cat("✅ Tables saved to outputs/tables \n")
```

## 🔁 Part 9 — Reproducibility Session Information

```{r}
#| label: session-info
#| code-fold: true

sessionInfo()
```

::: callout-tip
## Saving your renv snapshot

After the workshop, lock your package versions so results are reproducible:

```{r}
#| eval: false

renv::snapshot()
```

Then commit the updated `renv.lock` to GitHub:

``` bash
git add renv.lock
git commit -m "update renv snapshot after workshop"
git push
```
:::

## 📚 Summary

| Step | What we did |
|------------------------------------|------------------------------------|
| **Load data** | Read counts matrix and sample metadata |
| **Subset** | Kept only normal-condition samples for a clean 2-group comparison |
| **Pre-filter** | Removed genes with very low counts |
| **DESeq2** | Ran normalisation, dispersion estimation and Wald test |
| **QC** | PCA and sample distance heatmap |
| **Results** | Extracted Shank3 vs WT comparison with LFC shrinkage |
| **Visualize** | MA plot, volcano plot, top gene counts |
| **Export** | Saved tables to `outputs/tables/` and plots to `outputs/figures/` |

::: callout-note
## 🔭 Next Steps (beyond this workshop)

- Gene Ontology (GO) enrichment analysis with `clusterProfiler` and `fgsea`
- Heatmap of top DEGs with `pheatmap`
- More complex designs: adding condition and interaction terms
:::

------------------------------------------------------------------------

Next up: [**Apptainer**](08-apptainer.qmd) — running this same pipeline inside a
container, so the environment travels with the code.
